Discrete Ordinates

### 2017-06-12 docprep/chapmeth.tex
Updates

The $S_N$ quadrature ensures 90$\deg$ symmetry about any axis. In order to ensure the rotational symmetry, the $S_N$ quadrature has a single degree of freedom. The $LQ_N$ quadrature (which is the specific $S_N$ implemented in DOCTORS) uses the remaining degree of freedom to minimize error caused when integrating Legendre polynomials since that class of polynomials are used to represent anisotropy in scattering (see Sec.~\ref{sec:aniso}). For a more through discussion of quadrature generation, see \citet{ref:lewise}.

@@

@@

\section{Uncollided Solution Methods}\label{sec:uncol}
One of the earliest problems with the discrete ordinate methods is the presence of ray effect. Ray effect arises due to the restriction of particle transport to a set of discrete directions. To illustrate ray effect, consider a point source streaming particles isotropically into vacuum. Since no scatter or absorption takes place in vacuum conditions, the particles can only stream and particles must be conserved. However, if particles are transported along a single discrete direction only as illustrated in Fig.~\ref{fig:rayeffect_ex} ($\hat{\Omega} = <\sqrt{2}/2, \sqrt{2}/2>$ in Fig.~\ref{rayeffect_ex} and~\ref{fig:rayeffect_comp}) the phenomena observed in Fig.~\ref{fig:rayeffect_comp} is observed. As the discretization becomes more refined, the isotropic source becomes more heavily biased in the $\hat{\Omega}$ direction. This phenomena is called ray effect and has been thoroughly studied by many \citep{ref:mathewsk} \citep{ref:tencerj}.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/proj6}
        \caption{}
        \label{fig:proj6}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/proj12}
        \caption{}
        \label{fig:proj12}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/proj16}
        \caption{}
        \label{fig:proj16}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/proj24}
        \caption{}
        \label{fig:proj24}
    \end{subfigure}
    \caption{Different numbers of fan beam projections ($N$). (a) $N = 6$. (b) $N = 12$. (c) $N = 16$. (d) $N = 24$.}\label{fig:fanproj}
\end{figure}

@@

\subsection{Cone Source and Multi-cone Source}
Cone sources are identical to fan beam sources except for the rejection criteria used to determine whether or not a voxel is inside the beam. The rejection criteria is a function of only a single parameter, $\theta$, the maximum permissible angle between the center of the beam, $\boldsymbol{s}$, and a ray toward the point of interest, $\boldsymbol{c}$.

The angle between $\boldsymbol{s}$ and $\boldsymbol{c}$ is
\begin{equation}\label{eq:zetacos}
\cos(\zeta) = \frac{<S_x, S_y, S_z>}{\sqrt{S_x^2 + S_y^2 + S_z^2}} \cdot \frac{<C_x, C_y, C_z>}{\sqrt{C_x^2 + C_y^2 + C_z^2}}
\end{equation}
which, if $\boldsymbol{s}$ and $\boldsymbol{c}$ are unit vectors, simplifies to
\begin{equation}\label{eq:zetacos}
\cos(\zeta) = S_x C_x + S_y C_y + S_z C_z.
\end{equation}
Once the angle is computed, the condition
\begin{equation}\label{eq:zetacon}
|\zeta| < \theta
\end{equation}
is evaluated to determine whether the ray is accepted or rejected.

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/beamconexz}
        \caption{Cone xz}
        \label{fig:beamconexz}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/beamfanxz}
        \caption{Fan xz}
        \label{fig:beamfanxz}
    \end{subfigure}
    \caption{The beam shapes.}\label{fig:beamfancone}
\end{figure}

\section{Dose Computation}
Dose can be computed in either one of two ways. The first way is to compute the fluence distribution and multiply by a fluence-to-dose conversion factor from ICRP 116 [CITE]. The second way to compute dose is to compute the energy deposition in each cell. In both methods, transport of secondary particles is neglected and all energy deposition is assumed to be local.

@@

The publications of the ICRP include fluence-to-dose conversion factors for a reference phantom from various particle types and patient orientations. For CT systems, the rotationally symmetric (ROT) orientation is the most appropriate since the beam rotates about the patient. Figure~\ref{fig:icrp116_rotconv} shows the energy dependent fluence-to-dose conversions for the ROT orientation.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/icrp116_rotconv}
  \end{center}
  \caption{Fluence-to-dose conversion factors from ICRP 116 for photons.}
\label{fig:icrp116_rotconv}
\end{figure}

@@

The alternative way to compute the dose is directly from the local energy deposition. The total deposition, $\mathcal{E}_{T,i}$ for voxel $i$ is
\begin{equation}\label{eq:dept}
\mathcal{E}_{T,i} = \mathcal{E}_{a,i} + \mathcal{E}_{s,i}
\end{equation} 
where $\mathcal{E}_{a,i}$ is the energy deposition from absorption and $\mathcal{E}_{s,i}$ is the energy deposition from scatter. The absorption deposition is
\begin{equation}\label{eq:depa}
\mathcal{E}_{a,i} = \sum_{g=0}^{G-1} \Sigma_{a,i}^g \varphi_i^g V_i \bar{E}^g F_1
\end{equation}
where $\bar{E}$ is the average energy (in MeV) of the $g^{th}$ group and $F_1$ is a unit conversion factor that converts MeV to Joules. The average group energy is assumed to be the average between the two bounding energies of the group. Equation~\ref{eq:depa} assumes total local energy deposition in the voxel of interest. The energy deposition from scatter is
\begin{equation}\label{eq:deps}
\mathcal{E}_{s,i} = \sum_{g=0}^{G-1} \sum_{h=g+1}^{G-1} \left[ \Sigma_{s,i}^{g,h} \varphi_i^g V_i (\bar{E}^g-\bar{E}^h) F_1 \right]
\end{equation}
making the total deposition
\begin{equation}\label{eq:dept2}
\mathcal{E}_{T,i} = V_i F_1 \sum_{g=0}^{G-1} \varphi_i^g \left[\Sigma_{a,i}^g \bar{E}^g + \sum_{h=g+1}^{G-1} \left\{ \Sigma_{s,i}^{g,h}  (\bar{E}^g-\bar{E}^h) \right\} \right]
\end{equation}
The mass of the voxel, $M_i$ is then computed
\begin{equation}
M_i = \rho_i V_i F_2
\end{equation}
where $\rho_i$ is the density in g/cm$^3$ and $F_2$ is a conversion factor to convert grams to kilograms. The dose (in Gy) to the voxel is then
\begin{equation}\label{eq:dosedep}
D_i = \frac{\mathcal{E}_{T,i}}{M_i} = \frac{F_1}{\rho_i F_2} \sum_{g=0}^{G-1} \varphi_i^g \left[\Sigma_{a,i}^g \bar{E}^g + \sum_{h=g+1}^{G-1} \left\{ \Sigma_{s,i}^{g,h}  (\bar{E}^g-\bar{E}^h) \right\} \right].
\end{equation}

>>>>

The $S_N$ quadrature ensures 90 degree symmetry about any axis. In order to ensure the rotational symmetry, the $S_N$ quadrature has a single degree of freedom. The $LQ_N$ quadrature (which is the specific $S_N$ implemented in DOCTORS) uses the remaining degree of freedom to minimize error caused when integrating Legendre polynomials since that class of polynomials are used to represent anisotropy in scattering (see Sec.~\ref{sec:aniso}). For a more through discussion of quadrature generation, see \citet{ref:lewise}. Figure~\ref{fig:quad} shows the simples three quadratures implemented in DOCTORS.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figs/quad2}
        \caption{}
        \label{fig:quad2}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figs/quad4}
        \caption{}
        \label{fig:quad4}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figs/quad6}
        \caption{}
        \label{fig:quad6}
    \end{subfigure}
    \caption{Some various $S_N$ quadratures. The black dots indicate the location of a ray. (a) $S_2$ (b) $S_4$ (c) $S_6$ }\label{fig:quad}
\end{figure}

@@

INSERT WHY NOT USE FLUX MOMENTS

@@

An alternate anisotropy treatment is to use the Klein-Nishina formula~\citep{ref:shultisj}. The Klein-Nishina formula describes the scattering cross section as
\begin{equation}
\sigma(E, \mu) \propto q^3 + q\mu^2
\end{equation}
where
\begin{equation}
q(E, \mu) = \frac{1}{1 + (E/E_0)(1-\mu)}
\end{equation}

\section{Uncollided Solution Methods}\label{sec:uncol}
One of the earliest problems with the discrete ordinate methods is the presence of ray effect. Ray effect arises due to the restriction of particle transport to a set of discrete directions. To illustrate ray effect, consider a point source streaming particles isotropically into vacuum. Since no scatter or absorption takes place in vacuum conditions, the particles can only stream and particles must be conserved. However, if particles are transported along a single discrete direction only as illustrated in Fig.~\ref{fig:rayeffect_ex} ($\hat{\Omega} = <\sqrt{2}/2, \sqrt{2}/2>$ in Fig.~\ref{fig:rayeffect_ex} and~\ref{fig:rayeffect_comp}) the phenomena observed in Fig.~\ref{fig:rayeffect_comp} is observed. As the discretization becomes more refined, the isotropic source becomes more heavily biased in the $\hat{\Omega}$ direction. This phenomena is called ray effect and has been thoroughly studied by many \citep{ref:mathewsk} \citep{ref:tencerj}.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/proj6}
        \caption{}
        \label{fig:proj6}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/proj12}
        \caption{}
        \label{fig:proj12}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/proj16}
        \caption{}
        \label{fig:proj16}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/proj64}
        \caption{}
        \label{fig:proj24}
    \end{subfigure}
    \caption{Different numbers of fan beam projections ($N$). (a) $N = 6$. (b) $N = 12$. (c) $N = 16$. (d) $N = 64$.}\label{fig:fanproj}
\end{figure}

@@

\subsection{Cone Source and Multi-cone Source}
Cone sources are identical to fan beam sources except for the rejection criteria used to determine whether or not a voxel is inside the beam. The rejection criteria is a function of only a single parameter, $\theta$, the maximum permissible angle between the center of the beam, $\boldsymbol{s}$, and a ray toward the point of interest, $\boldsymbol{c}$. The angle between $\boldsymbol{s}$ and $\boldsymbol{c}$ is
\begin{equation}\label{eq:zetacos}
\cos(\zeta) = \frac{<S_x, S_y, S_z>}{\sqrt{S_x^2 + S_y^2 + S_z^2}} \cdot \frac{<C_x, C_y, C_z>}{\sqrt{C_x^2 + C_y^2 + C_z^2}}
\end{equation}
which, if $\boldsymbol{s}$ and $\boldsymbol{c}$ are unit vectors, simplifies to
\begin{equation}\label{eq:zetacos}
\cos(\zeta) = S_x C_x + S_y C_y + S_z C_z.
\end{equation}
Once the angle is computed, the condition
\begin{equation}\label{eq:zetacon}
|\zeta| < \theta
\end{equation}
is evaluated to determine whether the ray is accepted or rejected. Figure~\ref{fig:beamfancone} shows a comparison of the cone beam and fan beam shapes. The most noticable difference is in the $xz$ plne where the cone beam is circular, but the fan beam produces a rectangular cross section.

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/beamconexz}
        \caption{Cone xz}
        \label{fig:beamconexz}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/beamfanxz}
        \caption{Fan xz}
        \label{fig:beamfanxz}
    \end{subfigure}
    \caption{Comparison of a cone beam (a), (c), and (e) to a fan beam (b), (d), and (f). The cone beam is 20 degrees in both the $xy$ and $yz$ planes but the fant beam is different. The cross section of the cone beam is cicular (e), but the fan beam cross section is rectangular (f).}\label{fig:beamfancone}
\end{figure}

\section{Dose Computation}
Dose can be computed in either one of two ways. The first way is to compute the fluence distribution and multiply by a fluence-to-dose conversion factor from ICRP~116~\citep{ref:icrp116}. The second way to compute dose is to compute the energy deposition in each cell. In both methods, transport of secondary particles is neglected and all energy deposition is assumed to be local.

@@

The publications of the ICRP include fluence-to-dose conversion factors for a reference phantom from various particle types and patient orientations. For CT systems, the rotationally symmetric (ROT) orientation is the most appropriate since the beam rotates about the patient. Figure~\ref{fig:icrp116_rotconv} shows the energy dependent fluence-to-dose conversions for the ROT orientation. ICRP 116 distributes the conversion factors as point-wise data which must be converted to group data. This is performed using linear interpolation between the ICRP 116 data points.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/icrp116_rotconv}
  \end{center}
  \caption{Fluence-to-dose conversion factors from ICRP 116 for photons and the group averaged conversion factors for the 47-group data library in DOCTORS.}
\label{fig:icrp116_rotconv}
\end{figure}

@@

The alternative way to compute the dose is directly from the local energy deposition. The total deposition, $\mathcal{E}_{T,i}$ for voxel $i$ is
\begin{equation}\label{eq:dept}
\mathcal{E}_{T,i} = \mathcal{E}_{a,i} + \mathcal{E}_{s,i}
\end{equation} 
where $\mathcal{E}_{a,i}$ is the energy deposition from absorption and $\mathcal{E}_{s,i}$ is the energy deposition from scatter. The absorption deposition is
\begin{equation}\label{eq:depa}
\mathcal{E}_{a,i} = \sum_{g=0}^{G-1} \Sigma_{a,i}^g \varphi_i^g V_i \bar{E}^g F_1
\end{equation}
where $\bar{E}$ is the average energy (in MeV) of the $g^{th}$ group and $F_1$ is a unit conversion factor that converts MeV to Joules. The average group energy is assumed to be the average between the two bounding energies of the group. Equation~\ref{eq:depa} assumes total local energy deposition in the voxel of interest. The energy deposition from scatter is
\begin{equation}\label{eq:deps}
\mathcal{E}_{s,i} = \sum_{g=0}^{G-1} \sum_{h=g+1}^{G-1} \left[ \Sigma_{s,i}^{g,h} \varphi_i^g V_i (\bar{E}^g-\bar{E}^h) F_1 \right]
\end{equation}
making the total deposition
\begin{equation}\label{eq:dept2}
\mathcal{E}_{T,i} = V_i F_1 \sum_{g=0}^{G-1} \varphi_i^g \left[\Sigma_{a,i}^g \bar{E}^g + \sum_{h=g+1}^{G-1} \left\{ \Sigma_{s,i}^{g,h}  (\bar{E}^g-\bar{E}^h) \right\} \right]
\end{equation}
The mass of the voxel, $M_i$ is then computed
\begin{equation}
M_i = \rho_i V_i F_2
\end{equation}
where $\rho_i$ is the density in g/cm$^3$ and $F_2$ is a conversion factor to convert grams to kilograms. The absorbed dose (in Gy) to the voxel is then
\begin{equation}\label{eq:dosedep}
D_i = \frac{\mathcal{E}_{T,i}}{M_i} = \frac{F_1}{\rho_i F_2} \sum_{g=0}^{G-1} \varphi_i^g \left[\Sigma_{a,i}^g \bar{E}^g + \sum_{h=g+1}^{G-1} \left\{ \Sigma_{s,i}^{g,h}  (\bar{E}^g-\bar{E}^h) \right\} \right].
\end{equation}
